---
title: "Ch3: Clean Data in R"
author: "T.C. Hsieh"
date: "Friday, January 30, 2015"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(comment="", message=FALSE, warning=FALSE)
```

## 3. Clean Data in R
### 3.1 Reshape data
- 資料的轉置 `t`
```{r}
call <- read.csv("data/OutgoingCalls.csv")
call

call.t <- t(call) # 資料轉置後，物件類別會變成 "matrix"
colnames(call.t) <- as.character(call.t[1,]) # 用第一列 (first row) 來對欄位命名
call.t <- call.t[-1,] # 移除第一列
call.t <- as.data.frame(call.t) # 把 "matrix" 轉回 "data.frame"
call.t
```

- Wide format to long format
許多好用的分析、繪圖函數：`ggplot2`, `lm`, `glm`, `gam` 需要long-format。
```{r}
library(reshape2)
library(ggplot2)
df <- melt(call, id.vars = "Name", variable.name = "Week", value.name="OutCalls")
df
ggplot(df, aes(x=Week, y=OutCalls)) + geom_bar(stat="identity")
ggplot(df, aes(x=Week, y=OutCalls, fill=Name)) + geom_bar(stat="identity", position="dodge")
```

- long to wide
```{r}
dcast(df, Name~Week, value.var="OutCalls") # call

dcast(df, Week~Name, value.var="OutCalls") # call.t
```


### 3.2. 讀取 YouBike sample data
- 請[到DSP資料中心](data.dsp.im)下載 [data](http://ckan.nispc.tw/dataset/youbike-small-data)
- 將csv檔讀入R的環境中 
```{r}
ubike <- read.csv("data/ubike-sample-data-utf8.csv") # For Mac user
# ubike <- read.csv("data/ubike-sample-data-big5.csv") # For Windows user
head(ubike)
```
- Troubleshooting
[欄位說明文件](http://www.dot.gov.taipei/public/mmo/dot/YouBike%E5%BE%AE%E7%AC%91%E5%96%AE%E8%BB%8A%E4%BB%8B%E6%8E%A5%E8%AA%AA%E6%98%8E%E6%96%87%E4%BB%B6.pdf)

### 3.3. Arramge Observations (Rows)
```{r}
library(dplyr)
arrange(ubike, tot) # 由小排到大

arrange(ubike, desc(tot)) # 由大排到小
```

### 小挑戰
試試用`arrange`對行政區域 (`sarea`) 排序會如何
```{r echo=FALSE, eval=FALSE}
arrange(ubike, sarea) # 由小排到大
```


### 3.4. Subset Observations (Rows)
- 利用`filter`篩選資料
```{r}
ubike <- tbl_df(ubike) #  "tbl" class 是dplyr的物件，可以更方便展示資料
dim(ubike)
ubike <- distinct(ubike) # 移除重複的 rows
dim(ubike)

sample_n(ubike, 5, replace=TRUE) # 隨機抽5筆資料

slice(ubike, c(1,11,21,31,41)) # 自訂抽取的資料列

filter(ubike, sno==1) # 該場站大約每4-5分鐘記錄一筆資料
```

### 小挑戰
- 利用`filter` 選擇信義區的所有場站資料
- 篩選出經緯度介於：25.02387 < lat < 25.04815, 121.55575 < lng < 121.5838
Hint: R語言對於區間的判斷可用 25.02387 < lat < 25.04815
```{r echo=FALSE, eval=FALSE}
filter(ubike, sarea=="信義區")
filter(ubike, lat < 25.04815 & lat > 25.02387,  lng < 121.5838 & lng > 121.55575) 
```



### 3.5. Subset Variables (Columns)
- 利用`select`選取欄位
- Select columns by name
```{r}
# 選擇欄位
ubike.sub <- select(ubike, mday, sno, sarea, sna, tot, sbi, bemp, nbcnt, act, lat, lng)
# 也可以這樣做select
# ubike.sub <- select(ubike, c(mday, sno, sarea, sna, tot, sbi, bemp, nbcnt, act, lat, lng))
ubike.sub

# 反向選取(移除))欄位
select(ubike.sub, -mday, -sno)
# 也可以這樣做
# select(ubike.sub, -c(mday, sno))
```

- Select columns by helper function
```{r eval=FALSE}
# Select columns whose name matches a regular expression.
select(iris, matches(".t."))

# Select columns named x1, x2, x3, x4, x5.
select(iris, num_range("x", 1:5))

# Select columns whose names are in a group of names.
select(iris, one_of(c("Species", "Genus")))

# Select columns whose name starts with a character string. 
select(iris, starts_with("Sepal"))
```


### 3.6. Make New Variables
- Use `mutate` to compute and append one or more new columns
```{r}
ubike.sub <- mutate(ubike.sub, rate = sbi/tot) # rate = 目前車輛數/總空位數

ubike.sub <- mutate(ubike.sub, hour=strftime(mday, format="%H")) # hour = 小時 (00-23)

id <- c("南港區", "內湖區", "大安區","松山區","信義區","文山區", "中正區","萬華區","士林區","北投區","中山區","大同區")
ubike.sub <- mutate(ubike.sub, city=ifelse(sarea%in%id, "台北市", "新北市")) # 標記場站屬於台北市/新北市
ubike.sub
```

### 小挑戰 
- 對`ubike.sub`物件創建一個名為`min`的變數，表示場站記錄時間的分鐘
- Hint: 利用`strftime`取出`mday`變數的分鐘數, `%M`
```{r eval=FALSE, echo=FALSE}
mutate(ubike.sub, hour=strftime(mday, format="%M")) 
```

### 3.7. Group and Summarize Data
- Use `summarise` to summarise data into single row of values or each column
- 計算資料中場站目前空位數的平均數、標準差、最大、最小值
```{r}
summarise(ubike.sub, bemp.avg=mean(bemp), bemp.sd=sd(bemp), bemp.max=max(bemp), bemp.min=min(bemp))
```

- Use `group_by` to group data into rows with the same value of selected variable
- Compute separate summary row for each group
- 對行政區 (`sarea`) 做分組，計算該組別的目前空位數的平均數、標準差、最大、最小值
```{r}
summarise(group_by(ubike.sub, sarea), 
          bemp.avg=mean(bemp), bemp.sd=sd(bemp), bemp.max=max(bemp), bemp.min=min(bemp))
```

### 小挑戰
- 依`sno`做分組，計算該組別的目前空位數的平均數、標準差、最大、最小值
- 依`hour`做分組，計算該組別的目前空位數的平均數、標準差、最大、最小值
- 依`sno`與`hour`做分組，計算該組別的目前空位數的平均數、標準差、最大、最小值
```{r eval=FALSE, echo=FALSE}
summarise(group_by(ubike.sub, sno), 
          bemp.avg=mean(bemp), bemp.sd=sd(bemp), bemp.max=max(bemp), bemp.min=min(bemp))

summarise(group_by(ubike.sub, hour), 
          bemp.avg=mean(bemp), bemp.sd=sd(bemp), bemp.max=max(bemp), bemp.min=min(bemp))

summarise(group_by(ubike.sub, sno, hour), 
          bemp.avg=mean(bemp), bemp.sd=sd(bemp), bemp.max=max(bemp), bemp.min=min(bemp))
```

### 3.8. Combine Data Sets
- Columns and rows binding
- `bind_cols(y, z)`, `bind_rows(y, z)`
```{r}
day <- data.frame(day=c("2015-12-08", "2015-12-08", "2015-12-08","2015-12-08"))
tab <- summarise(group_by(ubike.sub, hour), 
          bemp.avg=mean(bemp), bemp.sd=sd(bemp), bemp.max=max(bemp), bemp.min=min(bemp))
bind_cols(day, tab)
```

- Mutating Joins
```{r}
df1 <- summarise(group_by(ubike.sub, hour), 
          bemp.avg=mean(bemp), bemp.sd=sd(bemp), bemp.max=max(bemp), bemp.min=min(bemp))

df2 <- summarise(group_by(ubike.sub, hour), 
          sbi.avg=mean(sbi), sbi.sd=sd(sbi), sbi.max=max(sbi), sbi.min=min(sbi))

left_join(df1, df2, by="hour")
```

### 小挑戰
把`ubike.sub`依小時 (`hour`)、場站 (`sno`) 整理成以下資料表
```{r echo=FALSE}
tab <- summarise(group_by(ubike.sub, sno, hour), 
          sarea=unique(sarea)[1],
          sna=unique(sna)[1],
          lat=unique(lat)[1],
          lng=unique(lng)[1],
          tot=unique(tot)[1],
          bemp.avg=mean(bemp), bemp.sd=sd(bemp), bemp.max=max(bemp), bemp.min=min(bemp),
          sbi.avg=mean(sbi), sbi.sd=sd(sbi), sbi.max=max(sbi), sbi.min=min(sbi),
          nbcnt.max=max(nbcnt)
          )
head(data.frame(tab))
```



 